---
title: "Data Visualization: `ggplot` part 3"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce box plots using `ggplot.`  
2. Customize labels on axes using `labs` and `themes`.  
3. Use `color` and `fill` to customize plots and improve aesthetics.  

## Review
Now that you have been introduced to `ggplot`, let's review the plot types from last week and learn how to manipulate their aesthetics to better suit our needs. Aesthetics make a significant visual difference, but you can take it too far so remember that the goal is to produce clean plots that are not distracting.  

##Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the data
Let's revisit the mammal life history data to practice our ggplot skills. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*
```{r message=FALSE, warning=FALSE}
life_history <- read_csv("data/mammal_lifehistories_v2.csv", na="-999") %>% clean_names()
```

## Bar Plots
Bar plots count the number of observations in a categorical variable. What is the difference between `geom_bar` and `geom_col`?  

Make two bar plots showing the number of observations for each order using each geom type.  

Count the number of observations by order 

`geom_col`
```{r}
life_history %>% 
  count(order) %>% 
  ggplot(aes(x=order, y=n, fill = order))+
  geom_col()+ #specify an x and a y 
  coord_flip()
  
```

`geom_bar`
```{r}
life_history %>% 
  ggplot(aes(x=order, fill = order))+
  geom_bar()+ #its going to specify x and y for you, do not need to use count with geom bar, its doing the count for you 
  coord_flip()
```

Remember that ggplot builds plots in layers. These layers can significantly improve the appearance of the plot. What if we wanted a bar plot of the mean mass for each order?  

Make a table and plot below.  
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass_order = mean(mass, na.rm = TRUE))
```

```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass_order = mean(mass, na.rm = TRUE)) %>% 
  ggplot(aes(x=order, y=mean_mass_order))+
  geom_col()+ #using geom col because x and y are specified 
  coord_flip()
```

There are a few problems here. First, the y-axis is in scientific notation. We can fix this by adjusting the options for the session.
```{r}
options(scipen=999)#cancels scientific notation for the session
```

Next, the y-axis is not on a log scale. We can fix this by adding `scale_y_log10()`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass_order = mean(mass, na.rm = TRUE)) %>% 
  ggplot(aes(x=order, y=mean_mass_order))+
  geom_col()+ #using geom col because x and y are specified 
  coord_flip()+
  scale_y_log10()
```

Lastly, we can adjust the x-axis labels to make them more readable. We do this using `reorder`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass_order = mean(mass, na.rm = TRUE)) %>% 
  ggplot(aes(x=reorder(order, mean_mass_order), y=mean_mass_order))+ #reorder the x 
  geom_col()+ 
  coord_flip()+
  scale_y_log10()
```

## Scatterplots
Scatter plots allow for comparisons of two continuous variables. Make a scatterplot below that compares gestation time and weaning mass.
```{r}
life_history %>% 
  select(gestation, wean_mass) %>% 
  ggplot(aes(x=gestation, y=wean_mass))+
  geom_point()+ #comparison between two continuous variables 
  scale_y_log10()
```

## Boxplots
Box plots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range. Let's make a box plot that compares mass across orders.
```{r}
life_history %>% 
  ggplot(aes(x=order, y=mass, fill = order))+
  geom_boxplot()+ #for a range 
  coord_flip()+
  scale_y_log10()

```
```{r}

```

## Aesthetics: Labels
Now that we have practiced scatter plots, bar plots, and box plots we need to learn how to adjust their appearance to suit our needs. Let's start with labeling x and y axes.  

For this exercise, let's use the `ElephantsMF` data. These data are from Phyllis Lee, Stirling University, and are related to Lee, P., et al. (2013), "Enduring consequences of early experiences: 40-year effects on survival and success among African elephants (Loxodonta africana)," Biology Letters, 9: 20130011. [kaggle](https://www.kaggle.com/mostafaelseidy/elephantsmf).
```{r}
elephants <- read_csv("data/elephantsMF.csv") %>% clean_names()
```

Make a plot that compares age and height of elephants.
```{r}
elephants%>% 
  ggplot(aes(x=age, y=height))+
  geom_point()+
  geom_smooth(method = lm, se = F)
  
```

The plot looks clean, but it is incomplete. A reader unfamiliar with the data might have a difficult time interpreting the labels. To add custom labels, we use the `labs` command.
```{r}
elephants%>% 
  ggplot(aes(x=age, y=height))+
  geom_point()+
  geom_smooth(method = lm, se = F)+
  labs(title="Elephant Age vs Height", #use to customize the plot and make them clean and clear
       x="Age", 
       y="Height")
```

We can improve the plot further by adjusting the size and face of the text. We do this using `theme()`. The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` allows control of title position.
```{r}
elephants%>% 
  ggplot(aes(x=age, y=height))+
  geom_point()+
  geom_smooth(method = lm, se = F)+
  labs(title="Elephant Age vs Height", 
       x="Age", 
       y="Height")+
  theme(plot.title=element_text(size=rel(1.5), hjust=.5)) #title is 1.5x larger in font than the axis 
```

## Other Aesthetics
There are lots of options for aesthetics. An aesthetic can be assigned to either numeric or categorical data. `fill` is a common grouping option; notice that an appropriate key is displayed when you use one of these options.
```{r}
elephants%>% 
  ggplot(aes(x=sex, fill=sex))+
  geom_bar()+
  labs(title="Elephant Per Sex", 
       x="Sex", 
       y="Count")+
  theme(plot.title=element_text(size=rel(1.5), hjust=.5))
  
```

`size` adjusts the size of points relative to a continuous variable.
```
```
```{r}
life_history %>% 
  ggplot(aes(x=gestation, y=log10(mass), size=(mass)))+
  geom_point()

```

## That's it! Let's take a break and then move on to part 2!  

-->[Home](https://jmledford3115.github.io/datascibiol/)